---
title: "Solutions to WPA2"
---

# Basic R

## 12. Now it's your turn

**Task A** 

1. Create a numeric vector called `participants`, containing integer numbers from 1 to 20, using `c()` and `seq()`.
2. Create a character vector called `conditions`, of length 20, containing alternating values of "a" and "b" ("a", "b", "a", "b", "a", ...), using `rep()`.
3. Create a vector called `first_half`, containing only the first half of the `participants` vector's values.
4. Check that both `participants` and `conditions` have length 20, and that `first_half` has length 10, using `length()`.
5. Instead of the fifth element in `conditions`, insert a missing value. Print the `conditions` after the change.
6. Create a vector called `participants_cond`, by pasting together (using `paste()`) `participants` and `conditions`, separated by `_`. So, the first element should be "1_a". 
7. Check that the 5th element of participants_cond is still a missing value.

```{r}
# task A1
participants = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20)
participants = seq(from=1, to=20)

# task A2
x = c("a", "b")
conditions = rep(x, times=10, each=1)

# task A3
first_half = participants[1:(length(participants)/2)]
print(first_half)

# task A4
length(participants) == 20 & length(conditions) == 20
length(first_half) == 10

# task A5
conditions[5] = NA
print(conditions)

# task A6
participants_cond = paste(participants, conditions, sep="_")
print(participants_cond)

# task A7
is.na(participants_cond[5])
participants_cond[5] = NA
print(participants_cond)
```

**Task B**

1. Create a data frame called `my_data`, with as columns the vectors`participants`, `conditions` and `participants_cond` that you created before. Print the data frame to have a look if it worked.
2. Add a column called `response_times` made of 20 samples from the normal distribution, with mean .8 and standard deviation 1. Print the data frame to have a look if it worked.
3. Select the values of the `response_times` column that are negative and set them to 0. Print the data frame to have a look if it worked.
4. Create a new column, called `log_response_times`, made of the logarithm of `response_times`.
5. Add a column called `correct_response` made of 20 samples from the binomial distribution, with size 1 and probability of success .65. Print the data frame to have a look if it worked.
6. Calculate the mean proportion of correct responses and the mean response time.
7. Create two data frames, `data_correct` and `data_incorrect` made of, respectively, the subset of `my_data` where `correct_response` is 1, and the subset of `my_data` where `correct_response` is 0. Print the data frame to have a look if it worked. Print the result to check.

```{r}
# task B1
my_data = data.frame(participants, conditions, participants_cond)
print(my_data)
```

```{r}
# task B2
my_data$response_times = rnorm(n=20, mean=.8, sd=1)
print(my_data)
```

```{r}
# task B3
my_data[my_data$response_times < 0, "response_times"] = 0
print(my_data)
```

```{r}
# task B4
my_data$log_response_times = log(my_data$response_times)
print(my_data)
```

```{r}
# task B5
my_data$correct_response = rbinom(n=20, size=1, prob=.65)
print(my_data)
```

```{r}
# task B6
mean(my_data$response_times)
mean(my_data$correct_response)
```

```{r}
# task B7
data_correct = my_data[my_data$correct_response == 1,]
data_incorrect = my_data[my_data$correct_response == 0,]
print(data_correct)
print(data_incorrect)
```